Expert-In-The-Loop Causal Discovery: Integrating Expert Knowledge

### 2024-08-05 paper/main.tex
Updates

Due to these challenges, even when using automated algorithms, we often have to
rely on manual post hoc testing and modifications to the model learned by the
algorithm. Whether constructing the model manually or using an
algorithm, we must rely on domain expertise to construct or modify the
model. Hence, it is crucial to have tools to guide this modification process.
One of the ways to guide this modification is to use Conditional Independence
(CI) based model testing \cite{Ankan2021}. As DAGs imply CI statements that can
be tested in the given data using statistical CI tests, we can verify whether
the implied CIs are consistent in the data or not. If some of implied CIs are
not valid in the given dataset, we can modify the model accordingly.
Figure~\ref{fig:ci_testing} shows an example of model testing using CIs.
However, this CI testing approach has a few drawbacks. Firstly, as CIs are
implied only by missing edges in the model, we can not test whether an
incorrect edge is already present in the model. Secondly, the number of implied
CIs can be quite large. For example, in the moderately sized alarm network with 
$ 37 $ node, there are $287$ implied CIs. If we rely only on the p-values of the
CI tests, it can be difficult to figure which test to focus on.

On the other hand, in Structural Equation Models (SEMs) literature, more
emphasis has been placed on developing tools to help researchers manually
construct models while integrating their domain expertise. Researchers
typically develop an initial model based on domain knowledge and use various
tools to guide modifications to improve the fit of their model to data. This
model construction process is known as \emph{Specification Search}. A common
method for specification search is using \emph{modification indices}, which
identify potential changes to the model structure that improve the model fit.
These potential changes are then ranked by the ones that improve the fit the
most. This ranking allows the researchers to focus first on the suggestions
that can improve the model the most and while integrating their expertise to
choose the most optimal modifications.

@@

In this section, we develop a novel partial measure of association for mixed
data by extending the idea of partial correlation to mixed data. We do this by
combining a mixed-data residualization method with a multivariate measure of
association. Given $ n $ samples $ D = (x, y, \bm{z}) $ of variables $ X $, $ Y
$, and $ \bm{Z} $, we are interested in computing an estimate of the
conditional association $ \rho(X, Y; \bm{Z}) $ between $ X $ and $ Y $
conditioned on $ \bm{Z} $. As the residualization approach requires atleast one
conditional variable to be present, if $ \bm{Z} = \emptyset $, we use a
constant value for $ Z $. Similar to partial correlation, we first compute the
residuals for $ X $ and $ Y $ using $ \bm{Z} $ as the predictor variables. We
use the mixed data residualization approach from \citet{Ankan2023}. Depending
on the type of dependent variable, the residuals are defined as follows:

@@

Using this mixed data partial association measure, we propose a method to rank
CI tests, and use it for model construction and modification. If the user
desires to construct the model manually, they can also start with an empty
graph, or start with a DAG outputted by a causal discovery algorithm. Given a
DAG $ G $ and a dataset $ D $, for each pair for variable $ X $ and $ Y $ in $
G $ that does not have an edge between them, we compute the partial association
measure: $ \rho(X, Y; pa_G(X) \cup pa_G(Y)) $. The measure of association in
this case can be interpreted as the unexplained observed association between
them given the current model. We can then use this to rank pair of variables
which have very high unexplained association between them. Based on the domain
expertise, we can then choose one of these suggested pair of variables, decide
the direction of the edge between the variables, and add it to the model. Once
the new edge is added we can recompute the association of all other variables
and $ X $ and $ Y $.

>>>>

Due to these challenges, even when using automated algorithms, we often have to
rely on manual post hoc testing and modifications to the model learned by the
algorithm. Whether constructing the model manually or using an algorithm, we
must rely on domain expertise to construct or modify the model. Hence, it is
crucial to have tools to guide this modification process. One of the ways to
guide this modification is to use Conditional Independence (CI) based model
testing \cite{Ankan2023}. As DAGs imply CI statements that can be tested in the
given data using statistical CI tests, we can verify whether the implied CIs
are consistent in the data or not. If some of implied CIs are not valid in the
given dataset, we can modify the model accordingly. Figure~\ref{fig:ci_testing}
shows an example of model testing using CIs. However, using p-values of the CI
tests has a few drawbacks. Firstly, as CIs are implied only by missing edges in
the model, we can not test whether an incorrect edge is already present in the
model. For examples, in high sample size scenarios, we might get a significant
association between a pair of variables with very small association. Secondly,
the number of implied CIs can be quite large. For example, in the moderately
sized alarm network with $ 37 $ node, there are $287$ implied CIs. If we rely
only on the p-values of the CI tests, it can be difficult to figure which test
to focus on.

To deal with both this issues, in this paper, we develop a measure of partial
association which can be using in combination of the CI test to rank the
violations in these CI tests. On the other hand, in Structural Equation Models
(SEMs) literature, more emphasis has been placed on developing tools to help
researchers manually construct models while integrating their domain expertise.
Researchers typically develop an initial model based on domain knowledge and
use various tools to guide modifications to improve the fit of their model to
data. This model construction process is known as \emph{Specification Search}.
A common method for specification search is using \emph{modification indices},
which identify potential changes to the model structure that improve the model
fit. These potential changes are then ranked by the ones that improve the fit
the most. This ranking allows the researchers to focus first on the suggestions
that can improve the model the most and while integrating their expertise to
choose the most optimal modifications.

@@

In this section, we develop a novel partial measure of association for mixed
data by extending the idea of partial correlation to mixed data. As partial
correlations require two things: 1) A residualization method, and 2) measure of
associations between the residuals. We generalize this to mixed data by combining a
mixed data residualization method \citep{Ankan2023} with a multivariate measure
of association based on Pillai's Trace. Given $ n $ samples $ D = (x, y,
\bm{z}) $ of variables $ X $, $ Y $, and $ \bm{Z} $, we are interested in
computing an estimate of the conditional association $ \rho(X, Y; \bm{Z}) $
between $ X $ and $ Y $ conditioned on $ \bm{Z} $. As the residualization
approach requires atleast one conditional variable to be present, if $ \bm{Z} =
\emptyset $, we use a constant value for $ Z $. Similar to partial correlation,
we first compute the residuals for $ X $ and $ Y $ using $ \bm{Z} $ as the
predictor variables. We use the mixed data residualization approach from
\citet{Ankan2023}. Depending on the type of dependent variable, the residuals
are defined as follows:

@@

Using this mixed data partial association measure, we propose a method to rank
CI tests, and use it for model construction and modification. Given a
potentially empty DAG $ G $ and a dataset $ D $, for each pair for variable $ X
$ and $ Y $ in $ G $:
\begin{itemize}
	\item If there is no edge between $ X $ and $ Y $, we compute p-value for the CI test $ X \ci Y \rvert pa_G(X) \cup pa_G(Y) $ and the partial association measure: $ \rho(X, Y; pa_G(X) \cup pa_G(Y)) $.
	\item If there is an edge between $ X $ and $ Y $, we compute p-value for the CI test $ X \ci Y \rvert pa_\underling{G}(X) \cup pa_{\underline{G}(Y)} $ and the partial association measure: $ \rho(X, Y; pa_{\underline{G}}(X) \cup pa_{\underline{G}}(Y) $.
\end{itemize}

The value of this partial measure of association can be interpreted as - given
the current structure of the model how much of the observed correlation between
the variables $ X $ and $ Y $ in data is not explained by the model. To modify
the model based on this, we can look at the pair of variables that have an edge
between them. If any of these pair have a p-value greater than partial
association below our chosen threshold, we can remove that edge as it is
redundant. In the case when there are

We can then use this to rank pair of variables which have very high unexplained
association between them. Based on the domain expertise, we can then choose one
of these suggested pair of variables, decide the direction of the edge between
the variables, and add it to the model. Once the new edge is added we can
recompute the association of all other variables and $ X $ and $ Y $.
